Mining of Keystroke and Mouse Dynamics to Increase the Engagement of Students with Programming Assignments

### 2018-04-20 paper.tex
test
Kernel
Revising related work

There has been very little research reported on the effectiveness of the use of
keyboard and mouse dynamics as a sensory channel for affective recognition, and
the few have not been focused on programming. The preliminary work by Zimmerman
et al. \cite{zimmermann2003affective} describes a method to correlate user’s interactions (keyboard and
mouse) with an emotional state, measuring different physiological parameters. The work of Vizer, Zhou \& Sears \cite{vizer2009automated} also
uses sensory data based on the time elapsed between each key press, the task
given to users was to write a free-text and used linguistic features in order to
recognize both physical and emotional stress. The results show a classification
accuracy of 62.5\% for physical stress and 75\% from emotional stress; authors
argue that these results are comparable with other approaches in affective
computing.

They also stress that their methods must be validated further in
other contexts. Moods may have an impact on programmer\'s performance according
to Khan, Hierons y Brinkman \cite{khan2007mood}. It may be possible to detect moods
on the basis of information regarding the programmer\'s use of the keyboard and
mouse, and to integrate them into development environments that can improve
programmer performance. They briefly describe a further experiment that could
use keyboard and mouse but only as future work. There are other studies about
the behaviour of programmers, which are not directly concerned with affective
states but are nevertheless important to their performance. Eye tracking in
computing education is proposed in the work Busjahn et al., \cite{busjahn2014eye}, and it is
also used for assessing learner´s comprehension of C++ and Python programs in
Turner et al., \cite{turner2014eye}. Blikstein \cite{blikstein2011using} proposed
an automated technique to
assess, analyse and visualize students learning computer programming. Blikstein
employs different quantitative techniques to extract students’ behaviours and
categorize them in terms of programming experience.

@@

Research works related to Keystroke Dynamics (KD) have been carried out by using either
fixed-texts or free-texts \cite{gunetti2005keystroke}.
KD performed on fixed-texts
involves the recognition of typing patterns when typing a pre-established fixed-
length text, e.g., a password. In the other case, free-text KD achieves the
recognition of typing patterns when typing a text of arbitrary-length, e.g., a
description of an item. However, as noted by Janakiraman y Sim \cite{janakiraman2007keystroke}, most of
the research regarding KD is done on fixed-text input, the reason being that
fixed-text KD usually yields better results than free-text KD. Yet, the authors
of this work share the opinion with Janakiraman, R., and Sim, T., that it would
 be more useful if KD can handle free text as well as fixed text, this is also a
 requirement if the text is a program.

@@

>>>>

There has been very little research reported on the effectiveness of
the use of keyboard and mouse dynamics as a sensory channel for
affective recognition, and the few have not been focused on
programming. The preliminary work by Zimmerman et
al. \cite{zimmermann2003affective} describes a method to correlate
user’s interactions (keyboard and mouse) with an emotional state,
measuring different physiological parameters. The work of Vizer, Zhou
and Sears \cite{vizer2009automated} also uses sensory data based on
the time elapsed between each key press, the task given to users was
to write a free-text and used linguistic features in order to
recognize both physical and emotional stress. The results show a
classification accuracy of 62.5\% for physical stress and 75\% from
emotional stress; authors argue that these results are comparable with
other approaches in affective computing.

They also stress that their methods must be validated further in
other contexts. Moods may have an impact on programmer\'s performance according
to Khan, Hierons and Brinkman \cite{khan2007mood}. It may be possible to detect moods
on the basis of information regarding the programmer\'s use of the keyboard and
mouse, and to integrate them into development environments that can improve
programmer performance. They briefly describe a further experiment that could
use keyboard and mouse but only as future work. There are other studies about
the behaviour of programmers, which are not directly concerned with affective
states but are nevertheless important to their performance. Eye tracking in
computing education is proposed in the work Busjahn et al., \cite{busjahn2014eye}, and it is
also used for assessing learner´s comprehension of C++ and Python programs in
Turner et al., \cite{turner2014eye}. Blikstein \cite{blikstein2011using} proposed
an automated technique to
assess, analyse and visualize students learning computer programming. Blikstein
employs different quantitative techniques to extract students’ behaviours and
categorize them in terms of programming experience.

@@

It is obviously necessary, in order to detect the moods, to prove that
there is an actual relationship between them and the keyboard
activity. In fact, Keystroke Dynamics (KD) has been investigated  by using either
fixed-texts or free-texts \cite{gunetti2005keystroke}.
KD performed on fixed-texts
involves the recognition of typing patterns when typing a pre-established fixed-
length text, e.g., a password. In the other case, free-text KD achieves the
recognition of typing patterns when typing a text of arbitrary-length, e.g., a
description of an item. However, as noted by Janakiraman and Sim
\cite{janakiraman2007keystroke}, most of 
the research regarding KD is done on fixed-text input, the reason being that
fixed-text KD usually yields better results than free-text KD. Yet, the authors
of this work share the opinion with Janakiraman, R., and Sim, T., that it would
 be more useful if KD can handle free text as well as fixed text, this is also a
 requirement if the text is a program. This is what we are actually
 using in this paper, since programs are created by the student and do
 not have to follow exactly the same pattern.

@@

\begin{table*}[!t]
\centering
\caption{Performance of classifiers for Flow/Engaged  with 10-fold cross validation. }
\label{tab_performance}
    \begin{tabular}{ | l | l | l | }
    \hline
    Classifier   &  Accuracy            & kappa                 \\
    \hline
    Na\"ive Bayes     & 72.52\%  & 0.339  \\
    \hline
    Deep Neural Network     & 78.29\%  & 0.419  \\
    \hline
    Random Forest      & 78.81\%  & 0.360  \\
    \hline
    Gradient Boosted Trees     & 78.95\%  & 0.415  \\
    \hline
    KNN      & 76.19\%  & 0.242  \\
    \hline
    Na\"ive Bayes Kernel     & 80.22\%  & 0.346 \\
    \hline
    J48     & 78.19 \%  & 0.433  \\
    \hline

    \end{tabular}
\end{table*}
